A No-Go Theorem for $\psi$-Ontic Models

### 2022-06-10 Articles/2021-HarriganSpekkens/HS-ShortNote.tex
HarriganSpekkens: review after Andrea

\begin{abstract}
In this note we give a simple no-go theorem that explains why $\psi$-ontic models cannot be expected to reproduce quantum mechanics. We will see that, using information theoretic considerations, the lack of overlap of epistemic state requires all states to be orthogonal, which openly contradicts quantum theory. This argument is a simplified version of the one we presented in a \href{https://arxiv.org/pdf/2201.11842.pdf}{previous paper}.
\end{abstract}

@@

Given the influence of the HS framework in quantum foundations, and in particular the prominent role played by $\psi-$ontic models in several arguments supporting the reality of the quantum state---many of them built upon \cite{PBR:2012}---we re-examine the definitions and features of such models in order to understand whether they constitute a sound basis from which to draw conclusions on the nature of $\psi$. Our analysis suggests that they are not.

Referring to this, in this note we provide a no-go theorem for $\psi-$ontic models which is a simplified version of the main result contained in our previous paper \cite{Carcassi:2022}. More precisely, here we show that, using information theoretic considerations, $\psi-$ontic models cannot reproduce QM since the lack of overlap of epistemic state requires all states to be orthogonal, which openly contradicts quantum theory. Therefore, we conclude, they should not be employed in order to classify quantum models.\footnote{We assume that the reader is familiar with HS definitions.}

@@

\textbf{Theorem:} \emph{Non-overlapping distributions can only represent orthogonal states}.

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping distributions over a space $\Lambda$, therefore eq. \ref{entropy_nonoverlap} applies. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
	H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.\ \ref{entropy_mixed}, it follows that $p$ must be zero. That is, we must have that
\begin{equation}\label{orthogonal}
	 \langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

Hence, the non-overlapping assumption built into the $\psi$-ontic model necessarily implies that all pure states are orthogonal. Since this is not true in quantum mechanics, any $\psi$-ontic model will fail to reproduce the results of quantum information, quantum statistical mechanics and, therefore, quantum theory in general.

In retrospect this should not be surprising for two reasons. First of all, the case where all states are orthogonal corresponds exactly to the case in which all observables commute, i.e.	\ the classical case. Therefore, we are simply finding that the implicit use of classical probability in $\psi$-ontic models pushes us to classical interpretation of quantum models. Secondly, we already know that quantum information theory has features that are not reproducible in classical information theory, and that is exactly what makes it such an exciting new field.

@@

As $\rho$ is in a diagonal form, the entropy is given by eq.~\ref{entropy}.\end{document}

>>>>

\begin{abstract}
In this note we give a simple no-go theorem that explains why $\psi$-ontic models cannot be expected to reproduce quantum mechanics. We will see that, using information theoretic considerations, the lack of overlap of epistemic states requires all states to be orthogonal, which openly contradicts quantum theory. This argument is a simplified version of the one we presented in a \href{https://arxiv.org/pdf/2201.11842.pdf}{previous paper}.
\end{abstract}

@@

Given the influence of the HS framework in quantum foundations, and in particular the prominent role played by $\psi$-ontic models in several arguments supporting the reality of the quantum state---many of them built upon \cite{PBR:2012}---we re-examine the definitions and features of such models in order to understand whether they constitute a sound basis from which to draw conclusions on the nature of $\psi$. Our analysis suggests that they are not.

Referring to this, in this note we provide a no-go theorem for $\psi$-ontic models which is a simplified version of the main result contained in our previous paper \cite{Carcassi:2022}. More precisely, here we show that, using information theoretic considerations, $\psi$-ontic models cannot reproduce QM since the lack of overlap of epistemic states requires all states to be orthogonal, which openly contradicts quantum theory. Therefore, we conclude, they should not be employed in order to classify quantum models.\footnote{We assume that the reader is familiar with HS definitions.}

@@

\emph{\textbf{Theorem: Since non-overlapping distributions can only represent orthogonal states, $\psi$-ontic models cannot be consistent with quantum theory.}}

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping distributions over a space $\Lambda$, therefore eq. \ref{entropy_nonoverlap} applies. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
	H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. That is, we must have that
\begin{equation}\label{orthogonal}
	 \langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

Hence, the non-overlapping assumption built into the $\psi$-ontic model necessarily implies that all pure states are orthogonal. Since this is not true in quantum mechanics, any $\psi$-ontic model will fail to reproduce the results of quantum information, quantum statistical mechanics and, therefore, quantum theory in general. $\square$

In retrospect this should not be surprising for two reasons. First of all, the case where all states are orthogonal corresponds exactly to the case in which all observables commute, i.e.~the classical case. Therefore, we are simply finding that the implicit use of classical probability in $\psi$-ontic models pushes us to classical interpretation of quantum models. Secondly, we already know that quantum information theory has features that are not reproducible in classical information theory, and that is exactly what makes it such an exciting new field.

@@

As $\rho$ is in a diagonal form, the entropy is given by eq.~\ref{entropy}.

\end{document}
